---
title: "Elderly Heat Relationships"
format: html
editor: visual
editor_options:
  markdown:
    wrap: sentence
execute:
  echo: false      # hides the code
  warning: false   # hides warnings
  message: false   # hides package-loading messages
  output: false    # hides printed output and plots
---

## Set up

### Install packages

Copy and paste to console:

if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman", repos = "https://cloud.r-project.org") pacman::p_load(tidyverse, sf, tigris, cowplot) options(tigris_use_cache = TRUE)

### Install libraries

```{r}
library(tidyverse)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)  # saves downloads so renders are fast
library(cowplot)
```

### Read data

```{r}
eh <- read.csv("data/elderlyheat.csv")
```

```{r}
names(eh)
```

## Wrangle

### Select only variable needed

```{r}
eh1 <- eh |> select(1, 2, 9, 29)
```

### Rename variables

```{r}
eh1 <- eh1 |> rename(id = 1, GEOID = 2, pop65 = 3, heatwaves = 4)
```

```{r}
names(eh1)
```

### Make GEOID text 11 characters (for join)

```{r}
eh1 <- eh1 |>
  mutate(GEOID = str_pad(as.character(GEOID), 11, pad = "0"))
```

```{r}
str(eh1)
```

## Retrieve County and Census tract polygons

```{r}
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)  # saves downloads so renders are fast

# County outline
county <- counties(state = "CA", cb = TRUE, year = 2023) |>
  filter(NAME == "Santa Clara")

# Census tracts in Santa Clara County (best for a pop65 x heatwaves map)
tracts_sc <- tracts(state = "CA", county = "Santa Clara", cb = TRUE, year = 2023)

plot(st_geometry(tracts_sc))
```

### Join data to census tracts

```{r}

# 1. Join your data (needs a GEOID column matching the tracts)
areas <- tracts_sc |>
  left_join(eh1, by = "GEOID")

# 2. Check the join worked: these counts should be near zero
sum(is.na(areas$pop65))
sum(is.na(areas$heatwaves))


```

## Create map and legend

```{r}
#use this code if quantile classificaion is desired.
#quantile(areas$pop65, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE)
#quantile(areas$heatwaves, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE)

```

```{r}
#| output: true
#| echo: true

library(tidyverse)
library(sf)
library(cowplot)


# 1. Classify with your own breaks (2 classes each)
areas <- areas |>
  mutate(
    pop65_c = cut(pop65, breaks = c(0, 1000, 2916),
                  labels = 1:2, include.lowest = TRUE),
    heat_c  = cut(heatwaves, breaks = c(57, 100, 111),
                  labels = 1:2, include.lowest = TRUE),
    bi_class = paste(pop65_c, heat_c, sep = "-")
  )

# 2. Palette: x = pop65 (blue), y = heatwaves (red), both high = dark purple
pal2 <- c(
  "1-1" = "#e8e8e8",   # low pop65, low heatwaves
  "2-1" = "#4a90c9",   # high pop65, low heatwaves (blue)
  "1-2" = "#e0605a",   # low pop65, high heatwaves (red)
  "2-2" = "#6a2c5a"    # high on both
)

# 3. Map
map <- ggplot(areas) +
  geom_sf(aes(fill = bi_class), color = "white", linewidth = 0.1,
          show.legend = FALSE) +
  scale_fill_manual(values = pal2, na.value = "grey80") +
  theme_void()

# 4. Diamond legend, drawn directly (no rotation needed)
tiles <- expand_grid(x = 1:2, y = 1:2) |>
  mutate(class = paste(x, y, sep = "-"))

# Four corner points of each tile, then rotated 45 degrees:
# X = x - y, Y = x + y
offs <- tibble(dx = c(-.5, .5, .5, -.5),
               dy = c(-.5, -.5, .5, .5),
               ord = 1:4)

diamond <- tiles |>
  cross_join(offs) |>
  mutate(X = (x + dx) - (y + dy),
         Y = (x + dx) + (y + dy)) |>
  arrange(class, ord)

legend <- ggplot() +
  geom_polygon(data = diamond, aes(X, Y, group = class, fill = class),
               color = "white", linewidth = 0.6) +
  scale_fill_manual(values = pal2, guide = "none") +
  
  # Corner labels (pop65, heatwaves)
  annotate("text", x = 0,     y = 0.55, label = "Low, Low",   size = 3, fontface = "bold") +
  annotate("text", x = 0,     y = 5.45, label = "High, High", size = 3, fontface = "bold") +
  annotate("text", x = 2.15,  y = 3,    label = "High, Low",  size = 3, fontface = "bold", hjust = 0) +
  annotate("text", x = -2.15, y = 3,    label = "Low, High",  size = 3, fontface = "bold", hjust = 1) +
  
  # Axis titles along the lower edges
  annotate("text", x = 1.5,  y = 1.5, label = "Population 65+",
           angle = 45,  size = 3.2) +
  annotate("text", x = -1.5, y = 1.5, label = "Heatwaves",
           angle = -45, size = 3.2) +
  
  coord_fixed(xlim = c(-4, 4), ylim = c(0, 6), clip = "off") +
  theme_void()

# 5. Combine
ggdraw() +
  draw_plot(map, 0, 0, 1, 1) +
  draw_plot(legend, x = 0.01, y = 0.01, width = 0.4, height = 0.4)
```

**How to read it**

- Each corner label reads (pop65, heatwaves).
  The top point is High, High (dark purple), the bottom is Low, Low (gray), the right is High, Low (blue), and the left is Low, High (red).

- “Population 65+” runs along the lower-right edge, the side that moves from low to high pop65.
  “Heatwaves” runs along the lower-left edge, which moves from low to high heatwaves.

**Adjustments**

- **More room for the titles:** move them outward by changing their `x`/`y`, for example `x = 1.7, y = 1.3` and `x = -1.7, y = 1.3`.
  Or widen the plot limits to `xlim = c(-4.5, 4.5)`.

- **Bigger or smaller legend:** change `width` and `height` in `draw_plot()`.
  Keep them equal for an 8x6 figure, which preserves the legend’s shape.

- **Text size:** adjust `size = 3` on the corner labels.
  ggplot sizes are in mm, so 3 is about 8.5 pt.

- **Position:** change `x` and `y` in `draw_plot()`.
  For example, `x = 0.58, y = 0.01` moves the legend to the lower right.

  #### Note:

- This uses `cross_join()`, which needs dplyr 1.1.0 or newer.
  If you get an error, run `update.packages("dplyr")`.
